Scalable Clustering of Genotype Information using MapReduce

### 2014-11-07 doc/publications/journal2014/paper.tex
pre-processing section

\subsection*{Pre-processing}
\label{Sec:preprocessing}
Mahout requires sequence files as input to its clustering algorithms, and it provides libraries for writing vectors to this file format.
Separate vectors for each sample, in this case, an individual, contain entries for each of the sample's features, where each entry is
a \texttt{double}. Although additional samples can subsequently be appended to sequence files, the entire feature-set for each sample must be
appended in a single write operation. Because the features for each sample are present in columns, as tab-separated elements in VCF
files, we recursively transpose the data into CSV files. Each comma-separated row contains the feature-set for a sample,
as well as the sample identity. Concurrently, we convert the values from the VCF file to doubles, as required for Mahout's sequence
file format and the many algorithms this format can serve as input for. We set the value of the double to the Hamming distance between
each variant and the reference genome. Therefore, a value identical to the reference will be 0, a heterozygous variant on either strand
will be 1 and a homozygous variant will be 2.
At this stage, the CSV files from more than one VCF file are concatenated into the one file. This resulting CSV file serves as the input for
a custom Java function. This iterates over the CSV file, extracting the sample name and feature-set from each line, and calls a Mahout
function to store this data as a named \texttt{DenseVector} object. The vectors are populated primarily by zeroes, so we convert this object to 
the more efficient \texttt{SequentialAccessSparseVector} object, which stores only non-zero doubles in an array, and maps their position in a second array.

This object is serialized, and the resulting file can serve as input to the majority of Mahout's algorithms.

>>>>

\subsection*{Pre-processing}
\label{Sec:preprocessing}
Mahout requires Hadoop's `SequenceFiles' as input to its machine learning algorithms. SequenceFiles are binary flat files consisting of key/value pairs. In this case, each of the keys is a
label for an individual. The associated value for each key is a `sparse vector' representing the individual's genomic variants. We use sparse vectors, rather than
dense vectors, to significantly reduce processing time and required disk space. Each sparse vector consists of two separate vectors. An item in the first vector is
effectively a key for an item at the same position in the send vector. The key represents a genomic location, where its value represents the variant. Zero values (positions where there is no variant)
are omitted. Therefore, the fewer variants in an individual's genome, the smaller the sparse vector size. In comparison, `dense vectors' are single arrays that
include every value. Therefore, the size of a dense vector isn't reduced with a large number of non-variants.

The values of the sparse vectors, which, as previously mentioned, represent an individuals genomic variants, are a \texttt{double} java type.
We set the value of each double to the Hamming distance between each variant and the reference genome. Therefore, a value identical to the reference will be 0 (and omitted from the sparse vector),
a heterozygous variant will be 1 and a homozygous variant will be 2.

We wrote a custom `MapReduce' class to convert VCF files to the aforementioned sequence files. This should work on any number of files,
of any file size; limited only by disk space and time. The `map' step creates key value pairs where the primary key is an individual, the secondary key is a variant location, and the value is the variant.
We used a secondary key as this allowed us to implement custom functions to collect and sort the output of each mapper, based on this secondary key. This significantly reduced
the overall processing time. The `reduce' step collects these sorted key/value pairs by primary key (individual ID) and adds the secondary key (location) and value (variant) to a sparse array. The algorithm
saves these sparse arrays and keys to disk in the aforementioned SequenceFile format.
